Changes in Patterns of Geometric Sign Groupings as a Model of Social Dynamics Across Aurignacian Phases

### 2025-06-09 paper/paper.qmd
stash

---
title: "Changes in Patterns of Geometric Sign Groupings as a Model of Social Dynamics Across Aurignacian Phases"
author: Sophie Cooper
format: docx
echo: false
execute: 
  warning: false
bibliography: references.bib
link-citations: true
---
# Introduction

The Upper Paleolithic period experienced a rise in more complex social organizations across Europe [@Boyd_Silk_2020], coupled with a biological transition from Neanderthals to early modern humans. Archaeologists have frequently relied upon variations in artwork, the large rise of which is a distinctive quality of the period, as a proxy for cultural difference in order to understand the shifting patterns of social organization and diversity of the Upper Paleolithic [@baker2024evidence; @Sauvet_2018; @Kuhn_Stiner_2007; @Vanhaeren_d; @Fuentes_Lucas_Robert_2019]. Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex (ca. 43 - 30 k BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], some of which are decorated with geometric signs [@dutkiewicz2020]. In this paper, we investigate the social organization dynamics by analyzing the distribution and variation of geometric sign types found on mobile objects from the Upper Paleolithic in Europe across four phases of the Aurignacian period.

@@

We hypothesize that these phases, and the environmental changes across the Aurignacian, may correspond to changes in social dynamics of cultural groups, as represented by patterns in the distribution of geometric signs. We aim to address the following questions: How do grouping dynamics of geometric signs change across the four phases of the Aurignacian? How do the groupings vary in diversity, size, connectivity, and distribution? Lastly, what do the patterns tell us about dynamics of social organization throughout the Aurignacian? We answer these questions by first identifying clusters within each phase using seriation and network analyses. We then test the strength of the clusters using perMANOVA tests. All of our programming in this study was conducted in RStudio.

@@

ggsave(maps_bar_plots,
      filename =  here::here("figures/map-and-hist-plot.png"),
       h = 10, w = 10,
       bg = "white")

knitr::include_graphics(here::here("figures/map-and-hist-plot.png"))

@@

  set.seed(500)
  jac <- vegdist(artifact_data %>% mutate_all(~ as.numeric(. > 0)), "jaccard", binary = TRUE)
  perman <- adonis2(jac~as.factor(group_data$group), 
                    method = "jaccard",
                    sqrt.dist = TRUE)
  perman <- perman %>% 
    data_frame() %>% 
    slice_head()
  perman_r <- round(perman$R2, 3)
  perman_p <- round(perman$`Pr(>F)`, 3)
  perman_f <- round(perman$F, 3)

@@

  final_table <- data_frame("PerMANOVA R2" = perman_r,
                            "PerMANOVA F" = perman_f,
                            "PerMANOVA p" = perman_p,
                            "Mantel R" = r_value_distance_correlation, 
                            "Mantel p" = p_value_distance_correlation)

@@

If the distribution of sign types can be explained by geographic distance, then we would see similarity in sign type distribution decreasing as geographic distance increases [@Herskind2023; @Rigaud_dErrico_Vanhaeren_2015]. This relationship can be investigated using the isolation-by-distance framework, which proposes that cultural difference can be primarily explained by geographic distance [@Lycett_2019]. To confirm that the distribution of sign types is influenced by cultural factors rather than geographic distance, we calculated the correlation between sign type distribution and geography for each phase. We used Mantel tests with a geographic distance matrix and a Jaccard similarity matrix, with the results shown in @tbl-mantel-table. None of the phases have a statistically significant correlation between sign type distribution and geographical distance, so the isolation-by-distance hypothesis is rejected.

@@

Seriation analysis is the arrangement of data into a linear order to reveal patterns [@Hahsler_Hornik_Buchta]. We used the Brower-Kile [-@Brower_Kile_1988] seriation algorithm from the seriation package [@seriation_pckge], in which a unidimensional sequence is generated by reordering the rows and columns to group the presences along the diagonal. To investigate connectivity among clusters of sites in the seriation solution, we used network analysis to represent and quantify the relationship between sites [@Mills_2017]. Each node is a site, and each edge is a measure of sign type distribution similarity, as computed by the Jaccard similarity algorithm. To avoid cluttering the graph, all edges below the value of 0.2 were removed. Results were plotted with the Fruchterman-Reingold layout [@Fruchterman_Reingold_1991], a force-directed graph layout for uniform edge lengths, and network analysis was performed using the vegan [@Vegan2001], statnet [@statnet_package] and igraph [@igraph_package] packages.

##  PerMANOVA for Cultural Connection Strength

PerMANOVA quantifies the level of variation between groups versus the level of variation within groups [@Anderson_2017; @Shennan_Crema_Kerig_2015]. The results are expressed by an R^2^ statistic, which quantifies how much of the variation in sign distribution can be explained by group membership, a pseudo-F statistic, and a p-value. The pseudo-F statistic compares the total sum of squared dissimilarities between groups to total sum of squared dissimilarities within groups. Larger pseudo-F statistics indicate more separation between groups. The p-value quantifies the probability of obtaining the observed difference if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance). The perMANOVA was performed using the vegan [@Vegan2001] and igraph [@igraph_package] packages.

@@

@@

In identifying the groups, we referenced both the seriation and network analyses. The results of the above tests allowed us to find two main groups of sites within each phase (@fig-group-map). In the Transitional phase, sites are widely spread out across Western Europe. This wider distribution may be a factor in weak connections seen in @fig-neighbor-net. In the Proto-Aurignacian, we see a slightly narrower distribution of sites westward but a spread northward, and more overlapping of sites between groups, which may be associated with the slightly stronger connectivity seen there. This phase is also associated with an increase in sites and sign types, which may be correlated with the increased proximity and population density. In the Early Aurignacian, we see a reduced distribution, with sites clustered closer together, which may be associated with the increased connectivity in that phase. In the Evolved Aurignacian, we see a wider distribution and more distance between sites, which may be associated with the reduced connectivity seen in that phase. The Mantel tests, as shown in @tbl-mantel-table, tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian has the highest correlation.

```{r}
#| label: tbl-strength-table
#| tbl-cap: "Table of the results per phase for the perMANOVA"
#| tbl-cap-location: bottom

strength_table <- full_table %>% 
  dplyr::select(`Aurignacian phase`, `PerMANOVA R2`, `PerMANOVA F`, `PerMANOVA p`)

## Grouping Dynamics Across the Four Phases of the Aurignacian
Our first aim was to determine how the grouping dynamics of geometric signs changed across four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and distribution. The results from @fig-seriation show that each phase can be divided into two main cultural groups, with sign type diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian. @fig-neighbor-net shows that group inter-connectivity and intra-connectivity varies from phase to phase, with weak intra-group connections and no inter-group connections in the Transitional phase, to stronger intraconnections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian. The perMANOVA tests show that the changes between the the Proto, Early, and Evolved Aurignacian phases are statistically significant, that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian has the least.

One consistent grouping dynamic throughout all four phases is the restricted/broader group divide, which may be influenced by variation in social structure. Despite the tendency to conceive of hunter-gatherers as homogeneous, previous studies have established variation in hunter-gatherer social structures [@Finlayson_Warren_2017; @Riches_1995; @Price_Brown_1985; @Lane_2017; @Pate_2006; @Johnson_2014; @Singh_Glowacki_2022; @Schwendler_2012], and additional studies have established how variation in ornamentation and other forms of material culture can be used to communicate information about social roles and status within groups [@Morwood_1987; @Sauvet_2018; @Mattson_2021; @Larsson_2006; @Moore_2010; @Alfonso-Durrruty_Giles_Misarti2015; @Vanhaeren_2005; @Conkey_1985; @Conkey_1978; @Tilley_1982; @Hodder1979; @Hodder1977; @Schwendler_2012]. It is therefore possible that, for some of the sites, the broader/restricted grouping indicates variation in social structure, with the broader groups having a greater variety of formalized social roles, and the restricted groups having fewer formalized social roles.

One interpretation of the connectivity seen in @fig-neighbor-net is as a representation of non-utilitarian mobility. Non-utilitarian mobility establishes regional social networks which facilitate the spread of information and the survival of hunter-gatherers in uncertain environments [@Whallon_2006; @Fitzhugh_Phillips_Gjesfeld_2011; @Lovis_Donahue_2011; @Hitchcock_Ebert_2011; @Romano_Lozano_Fern2002]. The exchange and production of decorative items is associated with non-utilitarian mobility and has been recognized as a sign of social contact in other regions and eras [@Whallon_2006; @Kelly_2024; @Kinahan_2013; @Newlander_2012; @Iizuka_Ferguson_Izuho_2022]. Previous studies have shown a range of variation in mobility among hunter-gatherer groups [@Binford_1990; @Kelly_1983; @Johnson_2014; @Padilla-Iglesias_Bischoff_2024; @Singh_Glowacki_2022]; some of the connections in @fig-neighbor-net may represent shorter-term information exchange, while others may indicate longer-term residential movement.

One form of longer-term residential movement is the aggregation/dispersion pattern, which is defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for subsistence or social reasons. Aggregation sites, associated with high degrees of connectivity, are the places where those group fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a more diverse artifact assemblage [@conkley_agg], inclusing, in some cases, decorative objects [@Kelly_2024]. Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994].

Aggregation activities may be identified by a higher than expected diversity index. @fig-div-plot plots the expected Shannon-Weaver diversity index [@Frerebeau_2025] for a given sample size against the actual diversity index per site [@Kintigh_1984]. The plot identifies multiple sites with a higher index than expected, such as Hohle Fels in the Transitional phase, Spy, La Ferrassie, and Les Cottés in the Proto-Aurignacian, Solutré, Hohle Fels, and Grottes de Fonds-de-Forêt in the Early Aurignacian, and Vogelherd, Bockstein-Törle, and Vindija Cave in the Evolved Aurignacian. Notably, most of these sites are not shown as having connections between groups in @fig-neighbor-net, except for La Ferrassie in the Proto-Aurignacian. This suggests a distinction between shorter-term information exchange (e.g. the highly interconnected sites in @fig-neighbor-net) and longer-term residential movement (e.g. the sites with higher than expected diversity in @fig-div-plot).

@@

This study adds to our knowledge of social networks within the Aurignacian period, as our results document dynamic patterns of social exchange and connection. Previous works [@bahn1982inter; @Horiuchi_Takakura_2019; @Rogers_2013; @baker2024evidence; @Golovanova_Doronichev_Doronicheva_Sapega_Shackley_2021; @Gamble_1982] have used material evidence analysis to reconstruct Upper Paleolithic social networks, which research has pointed to as a potential driver of cultural and evolutionary growth [@Greenbaum_Friesem_Hovers_Feldman_Kolodny_2019; @Creanza_Kolodny_Feldman_2017; @Cullen_1996; @Hill_Wood_Baggio_Hurtado_Boyd_2014; @Derex_Boyd_2016; @Romano_Lozano_Fern2022]. @Foley_Gamble_2009 places the emergence of complex inter-group connectivity at around 30 k BP, however, our results show evidence of inter-connectivity beginning at around 41 k BP, perhaps suggesting an earlier emergence of complex inter-connectivity than previously predicted. We attempted to rectify the low temporal resolution that tends to affect network analysis results when analyzing older, more time-averaged data sets [@Gravel-Miguel_Coward_2023] by dividing the data up into smaller time periods. Additionally, by focusing on change over time, we are able to see social networks as fluid and dynamic, rather than static and bounded [@Maher_Conkey_2019].

>>>>

---
title: "Changes in Patterns of Geometric Sign Groupings as a Model of Social Dynamics Across Aurignacian Phases"
author: Sophie Cooper
format: html
echo: false
execute: 
  warning: false
bibliography: references.bib
link-citations: true
---

# Introduction

The Upper Paleolithic period experienced a rise in more complex social organizations across Europe [@Boyd_Silk_2020], coupled with a biological transition from Neanderthals to early modern humans. Archaeologists have frequently relied upon variations in artwork, the large rise of which is a distinctive quality of the period, as a proxy for cultural difference in order to understand the shifting patterns of social organization and diversity of the Upper Paleolithic [@baker2024evidence; @Sauvet_2018; @Kuhn_Stiner_2007; @Vanhaeren_d]. Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex (ca. 43 - 30 k BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], some of which are decorated with geometric signs [@dutkiewicz2020]. In this paper, we investigate the social organization dynamics by analyzing the distribution and variation of geometric sign types found on mobile objects from the Upper Paleolithic in Europe across four phases of the Aurignacian period.

@@

We hypothesize that these phases, and the environmental changes across the Aurignacian, may correspond to changes in social dynamics of cultural groups, as represented by patterns in the distribution of geometric signs. We aim to address the following questions: How do grouping dynamics of geometric signs change across the four phases of the Aurignacian? How do the groupings vary in diversity, size, connectivity, and distribution? Lastly, what do the patterns tell us about dynamics of social organization throughout the Aurignacian? We answer these questions by first identifying clusters within each phase using seriation and network analyses. We then test the strength of the clusters using modularity and perMANOVA tests. All of our programming in this study was conducted in RStudio.

@@

ggsave(maps_bar_plots, 
      filename =  here::here("figures/map-and_hist-plot.png"),
       h = 10, w = 10,
       bg = "white")

knitr::include_graphics(here::here("figures/map-and_hist-plot.png"))

@@

  set.seed(500)
  jac <- vegdist(artifact_data %>% mutate_all(~ as.numeric(. > 0)), "jaccard", binary = TRUE)
  dm <- as.matrix(jac)
  modul_matrix <-
    graph_from_adjacency_matrix(1 - dm, mode = "undirected")
  mem <- as_membership(parse_number(group_data$group))
  as_group <- modularity(modul_matrix, membership = mem)
  as_group <- round(as_group, 3)

  perman <- adonis2(jac~as.factor(group_data$group), 
                    method = "jaccard",
                    sqrt.dist = TRUE)
  perman <- perman %>% 
    data_frame() %>% 
    slice_head()
  perman_r <- round(perman$R2, 3)
  perman_p <- round(perman$`Pr(>F)`, 3)
  perman_f <- round(perman$F, 3)

@@

  final_table <- data_frame("Modularity" = as_group, 
                            "PerMANOVA R2" = perman_r,
                            "PerMANOVA F" = perman_f,
                            "PerMANOVA p" = perman_p,
                            "Mantel R" = r_value_distance_correlation, 
                            "Mantel p" = p_value_distance_correlation)

@@

If the distribution of sign types can be solely explained by geographic distance, then we would see similarity in sign type distribution decreasing as geographic distance increases [. This relationship can be investigated using the isolation-by-distance framework, which proposes that cultural difference can be primarily explained by geographic distance [@Lycett_2019]. To confirm that the distribution of sign types is influenced by cultural factors rather than geographic distance, we calculated the correlation between sign type distribution and geography for each phase. We used Mantel tests with a geographic distance matrix and a Jaccard similarity matrix, with the results shown in @tbl-mantel-table. None of the phases have a statistically significant correlation between sign type distribution and geographical distance, so the isolation-by-distance hypothesis is rejected.

@@

Seriation analysis is the arrangement of data into a linear order to reveal patterns [@Hahsler_Hornik_Buchta]. We used the Brower-Kile [-@Brower_Kile_1988] seriation algorithm from the seriation package [@seriation_pckge], in which a unidimensional sequence is generated by reordering the rows and columns to group the presences along the diagonal. To investigate connectivity among clusters of sites in the seriation solution, we used network analysis to represent and quantify the relationship between sites [@Mills_2017]. Each node is a site, and each edge is a measure of sign type distribution similarity, as computed by the Jaccard similarity algorothm. To avoid cluttering the graph, all edges below the value of 0.2 were removed. Results were plotted with the Fruchterman-Reingold layout [@Fruchterman_Reingold_1991], a force-directed graph layout for uniform edge lengths, and network analysis was performed using the vegan [@Vegan2001], statnet [@statnet_package] and igraph [@igraph_package] packages.

## Modularity and PerMANOVA for Cultural Connection Strength

To evaluate the strength of the connections of the groups found in the seriation and network analyses, we used modularity and perMANOVA tests. The modularity test is a community structure network detection method, and quantifies the randomness of the groupings by measuring the fraction of the edges in the network that fall within groups, minus the expected value if the edges were placed at random [@Newman_2006; @Newman_Girvan_2004]. The results are expressed in a score of 0 to 1, with 0 meaning no significant subgroups, and 1 meaning perfect groupings. PerMANOVA quantifies the level of variation between groups versus the level of variation within groups [@Anderson_2017]. The results are expressed by an R^2^ statistic, which quantifies how much of the variation in sign distribution can be explained by group membership, a pseudo-F statistic, and a p-value. The pseudo-F statistic compares the total sum of squared dissimilarities between groups to total sum of squared dissimilarities within groups. Larger pseudo-F statistics indicate more separation between groups. The p-value quantifies the probability of obtaining the observed difference if the null hypothesis were true (we use an alpha value of 0.05 for the threshold of statistical significance). The modularity test and perMANOVA were performed using the vegan [@Vegan2001] and igraph [@igraph_package] packages.

@@

In identifying the groups, we referenced both the seriation and network analyses. The results of the above tests allowed us to find two main groups of sites within each phase (@fig-seriation).

```{r}
#| label: tbl-strength-table
#| tbl-cap: "Table of the results per phase for the modularity test and perMANOVA"
#| tbl-cap-location: bottom

## Modularity test

strength_table <- full_table %>% 
  dplyr::select(`Aurignacian phase`, Modularity, `PerMANOVA R2`, `PerMANOVA F`, `PerMANOVA p`)

proto_modularity <- proto_strength %>% 
  dplyr::select(Modularity)
evolved_modularity <- evolved_strength %>% 
  dplyr::select(Modularity)

The modularity score quantifies the strength of the groupings, showing that the variations in grouping strength between phases are minimal. The Proto-Aurignacian exhibits the strongest groupings, with a score of `r proto_modularity`, and the Evolved Aurignacian has the weakest groupings, with a score of `r evolved_modularity`. There is a slight increase in modularity between the Transitional period and the Proto-Aurignacian, before an even slighter decrease into the Early Aurignacian, followed by a larger increase with the Evolved Aurignacian.

## Grouping Dynamics Across the Four Phases of the Aurignacian

Our first aim was to determine how the grouping dynamics of geometric signs changed across four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and distribution. The results from @fig-seriation show that each phase can be divided into two main cultural groups, with sign type diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian. @fig-neighbor-net shows that group inter-connectivity and intra-connectivity varies from phase to phase, with weak intra-group connections and no inter-group connections in the Transitional phase, to stronger intraconnections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian. The modularity tests, as shown in @tbl-strength-table, show that the groupings are most distinct in the Transitional period and least distinct in the Early Aurignacian. The perMANOVA tests show that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian has the least. The Mantel tests, as shown in @tbl-mantel-table, tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian has the highest correlation.

@@

The geographic distribution of the groups change from phase to phase (@fig-group-map). In the Transitional phase, sites are widely spread out across Western Europe. This wider distribution may be a factor in weak connections seen in @fig-neighbor-net. In the Proto-Aurignacian, we see a slightly narrower distribution of sites westward but a spread northward, and more overlapping of sites between groups, which may be associated with the slightly stronger connectivity seen there. This phase is also associated with an increase in sites and sign types, which may be correlated with the increased proximity and population density. In the Early Aurignacian, we see a reduced distribution, with sites clustered closer together, which may be associated with the increased connectivity in that phase. In the Evolved Aurignacian, we see a wider distribution and more distance between sites, which may be associated with the reduced connectivity seen in that phase.

One factor that may influence the geographic structure of sites is the possibility that some sites functioned as aggregation sites. The aggregation/dispersion pattern is defined by @conkley_agg as a type of hunter-gather settlement pattern, in which groups both break off in dispersed fragments and congregate in large groups, typically for subsistence or social reasons. Aggregation sites, associated with high degrees of connectivity, are the places where those group fragments or individuals congregate, and some archaeological indicators of aggregation sites include proximity to certain environmental resources, as well as a diverse artifact assemblage [@conkley_agg]. Previous studies have also identified evidence of aggregation/dispersion patterns among the early modern humans of the Upper Paleolithic [@Maher_Conkey_2019; @bahn1982inter; @Bourdier_2013; @Svoboda_2013; @bahn1982inter; @white1992rethinking; @Soffer_Adovasio_Enloe_Audouze_Zubrow_2010; @montet-white1994].

A potential identifier of aggregation sites comes from @fig-div-plot, which uses the tabula package [@Frerebeau_2025] to generate the Shannon-Weaver diversity index of sign types per site, and plot the expected diversity index for a given sample size against the actual diversity index per site [@Kintigh_1984]. The plot identifies multiple sites that diverge from the diversity index predicted by sample size, including those with a higher index than expected, such as Hohle Fels in the Transitional phase, Spy, La Ferrassie, and Les Cottés in the Proto-Aurignacian, Solutré, Hohle Fels, and Grottes de Fonds-de-Forêt in the Early Aurignacian, and Vogelherd, Bockstein-Törle, and Vindija Cave in the Evolved Aurignacian. It is possible that those sites with a greater than expected diversity index acted as aggregation sites or served a similar purpose. La Ferrassie in the Proto-Aurignacian is also shown by @fig-neighbor-net to have multiple connections to sites in both groups, in addition to the diversity index shown in @fig-div-plot, perhaps suggesting that it served as a specifically inter-group aggregation site.

@@

This study adds to our knowledge of social networks within the Aurignacian period, as our results document dynamic patterns of social exchange and connection. Previous works [@bahn1982inter; @Horiuchi_Takakura_2019; @Rogers_2013; @baker2024evidence; @Golovanova_Doronichev_Doronicheva_Sapega_Shackley_2021; @Gamble_1982] have used material evidence analysis to reconstruct Upper Paleolithic social networks, which research has pointed to as a potential driver of cultural and evolutionary growth [@Greenbaum_Friesem_Hovers_Feldman_Kolodny_2019; @Creanza_Kolodny_Feldman_2017; @Cullen_1996]. @Foley_Gamble_2009 places the emergence of complex inter-group connectivity at around 30 k BP, however, our results show evidence of inter-connectivity beginning at around 41 k BP, perhaps suggesting an earlier emergence of complex inter-connectivity than previously predicted. We attempted to rectify the low temporal resolution that tends to affect network analysis results when analyzing older, more time-averaged data sets [@Gravel-Miguel_Coward_2023] by dividing the data up into smaller time periods. Additionally, by focusing on change over time, we are able to see social networks as fluid and dynamic, rather than static and bounded [@Maher_Conkey_2019].
